### 2019-02-07 ic-journal/expansion.tex
Synchronized htex with tex files; clear all highlights

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \emph{bisimulation} if, whenever
 $P\RR Q$, \DS{for all $\mu$} we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\enspace.
\end{enumerate}
 $P$ and $Q$ are \emph{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

We sometimes call bisimilarity the \emph{weak} one, to
distinguish it from \emph{strong} bisimilarity ($\sim$),
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums). For this, Milner introduced \emph{observational congruence}, also called \emph{rooted
  bisimilarity} \cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, if
\DS{for all $\mu$}:
%  \hl{for all $\mu\in \mathscr{L}\cup\{\tau\}$}
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

Besides reducing the rooted bisimiarity of two processes to
the bisimilarities of their  \DS{immediate derivatives}, the
above definition also brings up a proof technique \DS{for proving  rooted
bisimiarity results  by constructing a bisimulation.}
\begin{lemma}%{(Rooted bisimilarity by constructing a bisimulation)}
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
satisfies the following properties:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item the converse of (1) on the actions from $Q$.
\end{enumerate}
then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

>>>>

The equivalences we consider here are mainly \emph{weak} ones, in that they
abstract from the number of internal steps being performed:
\begin{definition}%[bisimilarity]
\label{d:wb}
A process relation ${\R}$ is a \emph{bisimulation} if, whenever
 $P\RR Q$, for all $\mu$ we have:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q \Arcap \mu Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$,implies that there is $P'$ such that $P \Arcap
  \mu P'$ and $P' \RR Q'$\enspace.
\end{enumerate}
 $P$ and $Q$ are \emph{bisimilar},
written as $P \wb Q$, if $P \RR Q$ for some bisimulation $\R$.
\end{definition}

We sometimes call bisimilarity the \emph{weak} one, to
distinguish it from \emph{strong} bisimilarity ($\sim$),
obtained by replacing in the above definition   the weak answer $
Q\Arcap\mu Q'$ with the strong  $Q \arr \mu Q'$.
Weak bisimilarity is not preserved by the sum operator (except for
guarded sums). For this, Milner introduced \emph{observational congruence}, also called \emph{rooted
  bisimilarity} \cite{Gorrieri:2015jt,Sangiorgi:2011ut}:
\begin{definition}%[rooted bisimilarity]
\label{d:rootedBisimilarity}
Two processes $P$ and $Q$ are \emph{rooted bisimilar}, written as $P
\rapprox Q$, if
for all $\mu$:
%  for all $\mu\in \mathscr{L}\cup\{\tau\}$
\begin{enumerate}
 \item  $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \wb Q'$;
 \item  $Q \arr\mu Q'$ implies that there is $P'$ such that $P
   \Arr\mu P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
\end{definition}

Besides reducing the rooted bisimiarity of two processes to
the bisimilarities of their immediate derivatives, the
above definition also brings up a proof technique for proving  rooted
bisimiarity results  by constructing a bisimulation.
\begin{lemma}%{(Rooted bisimilarity by constructing a bisimulation)}
\label{l:obsCongrByWeakBisim}
Given a (weak) bisimulation $\RR$, if two processes $P$ and $Q$
satisfies the following properties:
\begin{enumerate}
\item $P \arr\mu P'$ implies that there is $Q'$ such that $Q
   \Arr\mu Q'$ and $P' \RR Q'$;
\item the converse of (1) on the actions from $Q$.
\end{enumerate}
then $P$ and $Q$ are rooted bisimilar, i.e.~$P \approx^c Q$.
\end{lemma}

### 2019-02-07 ic-journal/formal.tex
Synchronized htex with tex files; clear all highlights

HOL4 is written in Standard ML, a single programming language which
plays three different roles:
%(The situation is quite different in other systems like Coq\footnote{\hl{The Coq Proof Assistant.} \url{https://coq.inria.fr}}.)
\begin{enumerate}
\item It serves as the underlying implementation language for the core HOL engine;
\item it is used to implement tactics (and tacticals) for writing proofs;
\item it is used as the command language of the HOL interactive shell.
\end{enumerate}
Moreover, using the same language HOL4 users can write complex automatic
verification tools by calling HOL's theorem proving
facilities. (The formal proofs of theorems in CCS theory
are mostly done by an \emph{interactive process} closely following
their informal proofs, with minimal automatic proof searching.)

@@

\emph{In this
formalisation we consider only single-variable equations/contractions.}
This considerably
 simplifies the required proofs in HOL, and  enhances the readability of
 proof scripts without loss of generality. For the  pencil-and-paper proofs, the
 multi-variable case is just a routine adaptation, \DS{while the   formalisation
 is not (this is also discussed  in Section~\ref{s:concl}).}

>>>>

HOL4 is written in Standard ML, a single programming language which
plays three different roles:
%(The situation is quite different in other systems like
%Coq\footnote{The Coq Proof Assistant. \url{https://coq.inria.fr}}).
\begin{enumerate}
\item It serves as the underlying implementation language for the core HOL engine;
\item it is used to implement tactics (and tacticals) for writing proofs;
\item it is used as the command language of the HOL interactive shell.
\end{enumerate}
Moreover, using the same language HOL4 users can write complex automatic
verification tools by calling HOL's theorem proving
facilities. (The formal proofs of theorems in CCS theory
are mostly done by an \emph{interactive process} closely following
their informal proofs, with minimal automatic proof searching.)

@@

\emph{In this
formalisation we consider only single-variable equations/contractions.}
This considerably
 simplifies the required proofs in HOL, and  enhances the readability of
 proof scripts without loss of generality. For the  pencil-and-paper proofs, the
 multi-variable case is just a routine adaptation, while the   formalisation
 is not (this is also discussed in Section~\ref{s:concl}).

### 2019-02-07 ic-journal/intro.tex
Synchronized htex with tex files; clear all highlights

\DS{
In Milner's versions of the unique-solution theorems for weak
equivalences,
 the  method has severe syntactical limitations:}
the equations must be \emph{(strongly) guarded and sequential}. That is, the
variables of the equations may only be used underneath a visible
prefix and proceed, in the syntax tree, only by the sum and prefix operators.
One way of overcoming such limitations is to replace the equations
 with special inequations called
\emph{contractions} \cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  The uniqueness of solutions of a system of contractions
is defined as with systems of equations: any two solutions must be \hl{(componentwise)} bisimilar.
The difference with equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in
\cite{sangiorgi2017equations} (in which the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of direct sums, due to the failure of the
substitutivity of contraction under direct sums.

The main goal of the work described in this paper is a rather
comprehensive formalisation of the core theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions
of \hl{equations and} contractions.
The formalisation, however, is not confined to the theory of unique
\DS{solutions}, but embraces a
significant portion the theory of CCS \cite{Mil89}, mostly because the
theory of unique solutions relies on a large number of more fundamental results.
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g.~the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g.~bisimulation up-to bisimilarity) \cite{Mil89,sangiorgi1992problem}, and the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy Lemma and Deng Lemma (Lemma 4.1 and 4.2 of
\cite{Gorrieri:2015jt}),
 and two long proofs saying \hl{that}  rooted bisimilarity is the coarsest (largest)
 congruence contained in (weak) bisimilarity: one following Milner's
 book \cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper \cite{van2005characterisation}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is an extensive experiment with the use of the HOL theorem prover with its
recent developments, including a package for coinductive definitions.

From the view of CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of \hl{contractions}, as
formally proving a previously known result gives us the
chance to see \emph{what is really needed} for establishing that result.
In particular, the existing theory \cite{sangiorgi2017equations} has
placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
We have thus refined the contraction-based proof technique, by moving to  
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.

Another \hl{benefit} of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures. 
Examples are: the results that rooted bisimilarity and rooted contraction are,
respectively, the coarsest congruence contained in weak bisimilarity 
and the coarsest precongruence contained in the contraction preorder; 
the result about unique solution of equations for weak bisimilarity that uses the
contraction preorder as an auxiliary relation, and other unique solution results (e.g.~%
the one for rooted in which
the auxiliary relation is rooted contraction); various forms of enhancements of the bisimulation
proof method (the `up-to' techniques).
In these cases, moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, \hl{eliminating} the risks 
of overlooking or missing details as in \hl{\emph{paper-and-pencil} proofs}.

>>>>

In Milner's versions of the unique-solution theorems for weak
equivalences,
 the  method has severe syntactical limitations:
the equations must be \emph{(strongly) guarded and sequential}. That is, the
variables of the equations may only be used underneath a visible
prefix and proceed, in the syntax tree, only by the sum and prefix operators.
One way of overcoming such limitations is to replace the equations
 with special inequations called
\emph{contractions} \cite{sangiorgi2015equations,sangiorgi2017equations}. Contraction is a
preorder that, roughly, places some efficiency
constraints on processes.  The uniqueness of solutions of a system of contractions
is defined as with systems of equations: any two solutions must be (componentwise) bisimilar.
The difference with equations is in the meaning of a solution:
in the case of contractions the solution is evaluated with respect to
the contraction preorder, rather than bisimilarity. 
With contractions, most syntactic limitations of the unique-solution theorem can be
removed.  One constraint that still remains in
\cite{sangiorgi2017equations} (in which the issue is bypassed using a more
restrictive CCS syntax)
is the occurrences of direct sums, due to the failure of the
substitutivity of contraction under direct sums.

The main goal of the work described in this paper is a rather
comprehensive formalisation of the core theory of CCS in the HOL
theorem prover (HOL4),  with a focus on the theory of unique solutions
of equations and contractions.
The formalisation, however, is not confined to the theory of unique
solutions, but embraces a
significant portion the theory of CCS \cite{Mil89}, mostly because the
theory of unique solutions relies on a large number of more fundamental results.
Indeed the formalisation encompasses the basic properties of strong and weak
bisimilarity (e.g.~the fixed-point and substitutivity properties), the
basic properties of
rooted bisimilarity (the congruence induced by weak
bisimilarity, also called observation congruence), and
their algebraic laws. Further extensions (beyond Nesi
\cite{Nesi:1992ve}) include four versions of ``bisimulation up to''
techniques (e.g.~bisimulation up-to bisimilarity) \cite{Mil89,sangiorgi1992problem}, and the
expansion and contraction preorder (two
efficiency-like refinements of weak bisimilarity). Concerning rooted bisimilarity, the formalisation
includes Hennessy Lemma and Deng Lemma (Lemma 4.1 and 4.2 of
\cite{Gorrieri:2015jt}),
 and two long proofs saying that  rooted bisimilarity is the coarsest (largest)
 congruence contained in (weak) bisimilarity: one following Milner's
 book \cite{Mil89}, with the hypothesis that no processes can use up
 all labels;
the other without such hypothesis, essentially formalising van Glabbeek's paper \cite{van2005characterisation}.
Similar theorems are proved for the rooted contraction preorder.
In this respect, the work is an extensive experiment with the use of the HOL theorem prover with its
recent developments, including a package for coinductive definitions.

From the view of CCS theory, this formalisation has offered us the possibility of
further refining the theory of unique solutions of contractions, as
formally proving a previously known result gives us the
chance to see \emph{what is really needed} for establishing that result.
In particular, the existing theory \cite{sangiorgi2017equations} has
placed limitations on the body of the contractions due to the
substitutivity problems of weak bisimilarity and other behavioural relations with respect
to the sum operator.
We have thus refined the contraction-based proof technique, by moving to  
\emph{rooted contraction}, that is, the coarsest precongruence contained in the contraction
preorder. The resulting unique-solution theorem is now valid for
\emph{rooted bisimilarity} (hence also for bisimilarity itself), and places no 
constraints on the occurrences of sums.

Another benefit of the formalisation is 
that we can take advantage of results about different 
equivalences and preorders that share similar proof structures. 
Examples are: the results that rooted bisimilarity and rooted contraction are,
respectively, the coarsest congruence contained in weak bisimilarity 
and the coarsest precongruence contained in the contraction preorder; 
the result about unique solution of equations for weak bisimilarity that uses the
contraction preorder as an auxiliary relation, and other unique solution results (e.g.~%
the one for rooted in which
the auxiliary relation is rooted contraction); various forms of enhancements of the bisimulation
proof method (the `up-to' techniques).
In these cases, moving between proofs there are only a few places in
which the HOL proof scripts have to be modified.
Then the successful termination of the proof gives us a guarantee that the proof is
complete and trustworthy, eliminating the risks 
of overlooking or missing details as in \emph{paper-and-pencil} proofs.
